From noise to knowledge: how randomness generates novel phenomena and reveals information

### 2018-02-21 paper/paper.Rmd
WIP

To the empricical ecologist, stochasticity, or more simply, "noise,"
is just that -- something which obscures patterns we are trying to infer [@Knape2011];
and an ever richer batteries of stastical methods are developed largely in an
attempt to strip away this undesirable randomness to reveal the patterns
beneath [@Coulson2001].  Over the past several decades, theoretical ecology has transitioned
from thinking of stochasticity in such terms; where noise is a nuciance
that obscures the "deterministic skeleton" of the underlying mechanisms,
to the recognition that stochasticity can itself be a mechanism for 
driving many interesting phenomena [@Coulson2004].  Yet this transistion from "noise the nuisance"
to "noise the creator" of ecological phenomena has had, with a few notable exceptions,
relatively little impact in empricial thinking about stochasticity.  One of the
most provocative of those exceptions has turned the classical notion of noise
the nuisance on its head: recognizing that noise driven phenomena can become a tool
to reveal underlying processes: to become "noise the informer."  Here I argue that
this third shift in perspective offers an opportunity to better bridge the divide
between respective primarily theoretical and primarily empricial communities by 
seeing noise not as mathematical curiousity or statistical bugbear, but as a
source for new opportunities for inference.  This shift in how we think about noise
is nonetheless easy to miss in the extensive literature

Origins of noise and motivation for studying stochastic models @Kendall1949 and @Bartlett1960

This trend from nuisance to creator
to informer is of course not the only or the most obvious trend in the 
extensive literature on stochastic modeling over the past few decades, but rather,
an instructive lens in which to draw out a lar

@Bjornstad2001 @Coulson2001 @Coulson2004

I will focus on examples that
illustrate this larger arc in how we are thinking about stochasticity as the role
it plays.

(e.g. see in particular
@Couslon2004, @Ovaskainen2010 on population extinction with 
a focus on environmental and colored noise, @Black2012 on intrinsic / demographic noise,

Few topics have recieved greater attention then the question of
the persistence of a given species [@Kendall1949;  @Bartlett1960; @Lande1993; @Saether1998; @Coulson2001; @Schreiber2010]

Excellent reviews and textbook introductions

@Bartlett1960 @Bjornstad2001 @Coulson2001 @Lande2003

See in particular reviews by @Ovaskainen2010 and @Schreiber2012

@Lee2017

The extensive literature on demographic and environmental noise has been
frequently reviewed elsewhere [@Black2012]

@Saether1997

I approach this synthesis in three parts.  First, I will sketch the ground-work
for the origins of noise in any model as statistical summary of large number of
additional processes from some lower level that are modeled only in aggregate.
This will provide the fundamentals we need to illustrate several key examples
of noise driven phenomena in ecological dynamics: noise the creator.  Finally,
we will be able to invert some of these examples; instead of going from process
to pattern, we will go from pattern to process: noise the informer.

I will anchor each of these sections using concrete examples from simple models.
It is easy to critique such simple models: after all, one might summarize decades
of research in noting the twin observations that (1)  ecology is complex and (2)
that complexity is important. In the stochasticty literature alone, much work 
has been done to understand and extend fundamental results to more complex models,
including the importance of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989;
@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g. @Durrett1994; @Dieckmann2000;
@Kerr2002; @Schreiber2010], Individual heterogeneity [e.g. @Vindenes2015; @Hart2016]
or the coexistence of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Schreiber2017]; 
correlated [@Ripa; @Ripa1999; @Petchey, @Schreiber2010; @Lee2017; @Spanio2017] or periodic
[@Coulson2001; @Bjornstad2001; @Keeling2001] structure in environmental noise, or
the interaction of ecological and evolutionary processes [@Saether1997; @Ozgul2009; @Lenormand2009; @Vindenes2015; @Schreiber2015],
not to mention the extensive literature in stochastic evolution and popluation genetics
[e.g. @Ewens2004, @Lewontin1969].

While acknowledging more realistic examples and more general results where available,
using specific and familiar simple models -- just complex enough to illustrate
the phenomenon but no more -- will provide us with a platform that is more accessible 
to both intution and exploration and extension either through mathematical manipulation or 
computer simulation.  In this spririt, I provide concise, detailed and commented
code for simulating each of the models presented in Appendix A.
Simulations permit poking and prodding of empirically-minded
investigation unencumbered by either experimental design or mathematical formalism.
Most stochastic models are expressed in the BUGS language, which
may be more familiar to empirical readers than corresponding
mathematical formulas. This also permits both efficient simulation
and potential estimation of parameters given sample data using the
R package, NIMBLE [@deValpine2017]; allowing the deductive models
illustrated here to readily function as inductive models with
parameters inferred from time-series data. A copy of this appendix
is maintained at <https://github.com/cboettig/noise-phenomena>, and
bug reports, suggestions or help requests are welcome through the
issue tracker.  A second appendix provides more mathematical details
for interested readers.  With these in place, we are ready to begin
our tour through three major areas of noisy phenomena: Origins of
noise, emergent phenomena, and noise-driven inference.

@@

@@

It is often suggested that stochasticity is "inherent" or "fundamental,"
e.g. "demographic stochasticity arises because the birth of an individual is probabilistic"
[@Melbourne2012]. It is easy to forget that this is merely an assertion about a 
model, not an assertion about reality.  Only in quantum mechanics do we find 
inherently probabilistic properties: births and deaths of organisms all have
far more mechanistic explanations which we simply summarize in statistical terms:
on average -- that is, summing over a wide range of possible mechanisms we have chosen
not to explicitly model -- an individual death occurs at rate $\lambda$.  
This is in fact an extrinisic factor (variables not modeled explicitly),
which makes birth appear probabilistic; (just as we describe the toss of a coin
as probabilistic when in fact it is a mechanistic outcome of Newton's laws).  
Demographic stochasticity arises
only subsequently, when we change scales from a description at the individual
level to one at the population level.

@Keeling2012; @Melbourne2012; @Ripa2012

This does not attempt to summarize all of the literature on the origins and
types of noise, which have been reviewed elsewhere [e.g. @Saether1997; @Coulson2004; @Ovaskainen2010; @Black2012; @Schreiber2012; @Lenormand2009]

justify simple models, 
citations adapting this to complex models:

Early
quantitative ecological models first formulated at the population
level as differential or difference equations: deterministic models
that seek to capture the essential features of a system while
neglecting any random component as mere noise [@Coulson2004].
Advances in both theoretical understanding and the advent of widely
available and efficient numerical methods have helped drive research
towards increasingly more explicit descriptions of how stochasticity
enters into quantitative models [@Black2012; @Ovaskainen2010].

@@

It is possible to simulate such a process exactly using Gillespie's
exact stochastic simulation algorithm [SSA; @Gillespie1977], which
simulates each individual birth or death as a separate event. Figure \ref{gillespie}
uses this algorithm to compare simulations of @Levins1969 patch model;
(see Appendix A for example code).  In the Levins'  model,
$n$ individuals compete for a
finite number of suitable habitats $N$.  Individuals die a constant
rate $e$, and produce offspring at a constant rate $c$ who then
have a probability of colonizing an open patch that is simply
proportional to the fraction of available patches, $1 - n/N$.

@@

@@

Stochasticity frequently enters our models in other ways as well.
For instance, most population-level models treat all individuals
within the population, or at least within a particular age or stage
class, as identical.  *Individual heterogeneity* in relevant traits
(growth rates, dispersal ability) can be represented probabilistically
as a distribution rather than a fixed value, making it function as
another form of random variation in a model [@Coulson2001;
@Schreiber2010; @Vindenes2015].  *Spatial heterogeneity* can likewise be reflected
in stochastic parameters of a spatially implicit model, or as
spatially varying environmental stochasticity in a spatially explicit
model.  As with other forms of noise [@Melbourne2008; @Bonsall2004],
interaction between individual or spatial variation and other sources
of noise can drive patterns including the potential for persistence,
coexistence, and dispersal of species [@Schreiber2010; @Hart2016;
@Schreiber2017].  *Observational error* acts
much like another form of stochasticity, arising from sampling
effects or limitations of instruments, etc. Yet unlike the sources
described above, this uncertainty does not impact the actual state
of our system, and thus does not give rise to any additional phenomena
or information about that process, though it can obscure it
[@Knape2011; @deValpine2002].

Other forms of uncertainty common in models are not stochastic
phenomena.  This includes any form of *reducible uncertainty*: In
constructing and estimating mathematical models of ecological
processes, we encounter uncertainty over our parameter estimates
and possible choice of models [@Ellison2004].  In some ways, this
uncertainty can act like the stochasticity described above: for
instance, both introduce variation into any forecast of future
outcomes. Yet unlike stochasticity, this uncertainty is reducible:
with additional observations, we can narrow the uncertainty over
parameters or candidate models. Thus we will not consider model and
parameter uncertainty, as they do not reflect intrinsic phenomena of 
the dynamics but only our own ignorance about those dynamics.  
*Deterministic chaos* [@Desharnais2001] is another mechanism of introducing
significant variation into a simple model. Chaos can a look a lot
like noise: indeed, it can be observationally indistinguishable
from a stochastic process [@Ellner2005; @Werndl2009].  Despite
substantial interest in previous decades and the relative ease of
encountering chaos in simple ecological models (particularly in
differences equations), and convincing reconstructions of chaotic
dynamics in laboratory manipulations [@Costantino1997] the prevalence
of chaotic dynamics as a driving factor in most ecological systems
remains unclear. After all, the great variation of natural populations
is hardly surprising: the undeniable influences of variation created
by all those things left out of our models, all that individual
detail from intrinsic, lower level processes and external environmental
processes that becomes aggregated in our simple models all but begs
for a stochastic description.

Much active research continues to explore the consequences of
introducing these more explicitly motivated and interacting sources
of noise. With some grounding in the origins of different types of
noise in place, we are ready to move to our second stop to see how
this noise can drive the creation of novel phenomena on a macroscopic
level.

@@

Few topics have recieved greater attention then the question of
the persistence of a given species [@Kendall1949;  @Bartlett1955; @Bartlett1960; @Lande1993; @Saether1998; @Coulson2001; @Schreiber2010]

and coexistence of multiple species

[@Caswell1978; @Tuljapurkar1980; @Chesson1981; @Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Schreiber2017]

Structured populations 
[@Cohen1979; @Tuljapurkar1989; @Metcalf2015; @Vindenes2017b]

See in particular reviews by @Ovaskainen2010 and @Schreiber2012

Individual heterogeneity 
@Vindenes2015; @Hart2016

@Lee2017

A natural focus of much research into stochastic ecological systems
has been on understanding the implications for persistence of
individual and coexistence of interacting species. Demographic noise
in particular has fundamental implications for the long-term
equilibrium: any finite population, however big, is guaranteed to
eventually go extinct under demographic noise [e.g. @Kendall1949; @Lande2003].  However, in most
models it is possible to define a notion of quasi-stationary 
distribution concentrated around the deterministic attractor,
and to prove the probability of extinction decreases exponentially with 
system size (e.g. see @Ovaskainen2010; @Schreiber2017 also reviews 
more recent literature which has established these results more generally). 
Generalizations about persistence and stable coexistence of interacting
species in stochastic models are difficult due to complex interactions
between demographic and environmental noise, population structure,
spatial heterogeneity, magnitude of noise, degree of auto-correlation and so forth
[e.g. @Higgins1997; @Coulson2001; @Melbourne2008; @Hart2016];
indeed care must be taken even in defining the appropriate notion of 
persistence [e.g. @Schreiber2006].  Instead of these long-term limits,
here we will focus on a few examples of stochastic phenomena 
that can emerge on shorter time scales and provide some intuition for 
how the interaction between stochastic an nonlinear dynamics can 
create these surprising results.

>>>>

To the empricical ecologist, stochasticity, or more simply, "noise,"
is just that -- something which obscures patterns we are trying to infer [@Knape2011];
and an ever richer batteries of stastical methods are developed largely in an
attempt to strip away this undesirable randomness to reveal the patterns
beneath [@Coulson2001].  Over the past several decades, theoretical ecology has transitioned
from thinking of stochasticity in such terms; where noise is a nuciance
that obscures the "deterministic skeleton" of the underlying mechanisms,
to the recognition that stochasticity can itself be a mechanism for 
driving many interesting phenomena [@Coulson2004].  Yet this transistion from "noise the nuisance"
to "noise the creator" of ecological phenomena has had, with a few notable exceptions,
relatively little impact in empricial thinking about stochasticity.  One of the
most provocative of those exceptions has turned the classical notion of noise
the nuisance on its head: recognizing that noise driven phenomena can become a tool
to reveal underlying processes: to become "noise the informer."  Here I argue that
this third shift in perspective offers an opportunity to better bridge the divide
between respective primarily theoretical and primarily empricial communities by 
seeing noise not as mathematical curiousity or statistical bugbear, but as a
source for new opportunities for inference.

As the literature on stochastic modeling in ecology is both extensive and often
technical, and so rather than attempt a comprehensive review, this synthesis
will focus on the simplest examples necessary to illustrate this broader trend
in the changing way we think about the role of noise and the opportunity
that shift can present to better bridge theoretical and empirical communities.
To that end, I will focus on examples from simple population models, echoing
the motivation of @Bartlett1960 to focus on basic theoretical aspects common
across models. In so doing, we will ignore the equally rich literature of 
stochastic models in evolution [@Lenormand2009] and population genetics
[e.g. @Ewens2004]. In an effort to make these themes accessible to more emprical
readers, I will also focus on simplest examples necessary, with only passing
mention of the rich literature extending these concepts to more realistic
models.  It is easy to critique this choice: indeed, one might summarize decades
of research in noting the twin observations that (1)  ecology is complex and (2)
that complexity is important. In the stochasticity literature alone, much work 
has been done to understand and extend fundamental results to more complex models,
including the importance of age and stage structure [e.g. @Cohen1979; @Tuljapurkar1989;
@Caswell2009; @Metcalf2015; @Vindenes2017b], spatial structure [e.g. @Durrett1994; @Dieckmann2000;
@Kerr2002; @Schreiber2010], Individual heterogeneity [e.g. @Vindenes2015; @Hart2016]
or the coexistence of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Schreiber2017]; 
correlated [@Ripa; @Ripa1999; @Petchey, @Schreiber2010; @Lee2017; @Spanio2017] or periodic
[@Coulson2001; @Bjornstad2001; @Keeling2001] structure in environmental noise, or
the interaction of ecological and evolutionary processes [@Saether1997; @Ozgul2009;
@Lenormand2009; @Vindenes2015; @Schreiber2015]. In this we follow @Levins1966 in
sacrificing realism for generality.

In this spririt, I provide concise, detailed and commented
code for simulating each of the models presented in Appendix A.
Simulations permit poking and prodding of empirically-minded
investigation unencumbered by either experimental design or mathematical formalism.
Most stochastic models are expressed in the BUGS language, which
may be more familiar to empirical readers than corresponding
mathematical formulas. This also permits both efficient simulation
and potential estimation of parameters given sample data using the
R package, NIMBLE [@deValpine2017]; allowing the deductive models
illustrated here to readily function as inductive models with
parameters inferred from time-series data. A copy of this appendix
is maintained at <https://github.com/cboettig/noise-phenomena>, and
bug reports, suggestions or help requests are welcome through the
issue tracker.  Appendix B provides more mathematical details
for interested readers.  With these in place, we are ready to begin
our tour through three major areas of noisy phenomena: Origins of
noise, emergent phenomena, and noise-driven inference.

<!--
I approach this synthesis in three parts.  First, I will sketch the ground-work
for the origins of noise in any model as statistical summary of large number of
additional processes from some lower level that are modeled only in aggregate.
This will provide the fundamentals we need to illustrate several key examples
of noise driven phenomena in ecological dynamics: noise the creator.  Finally,
we will be able to invert some of these examples; instead of going from process
to pattern, we will go from pattern to process: noise the informer.  
-->

@@

Many stochastic models are not explicit about whether the noise terms
they introduce are intended to reflect intrinsic or extrinsic factors.
Early quantitative ecological models first formulated at the population
level as differential or difference equations: deterministic models
that seek to capture the essential features of a system while
neglecting any random component as mere noise [@Coulson2004].
Advances in both theoretical understanding and the advent of widely
available and efficient numerical methods have helped drive research
towards increasingly more explicit descriptions of how stochasticity
enters into quantitative models.

@Black2012 provides an important 
review of intrinsic noise and approaches for addressing it, focusing
in particular on use of the van Kampen system size expansion as a
mathematical means to properly bridge between individual-based models
(IBMs, usually implemented only in simulations) and familiar population-level
dynamics (so called macroscopic dynamics).

; @Ovaskainen2010 the canonical on the highly influential @Leigh1981, @Lande1993, @Foley1994

This division between intrinsic or demographic noise and extrinsic or
environmental noise

@@

It is often suggested that stochasticity is "inherent" or "fundamental,"
e.g. "demographic stochasticity arises because the birth of an individual is probabilistic"
[@Melbourne2012]. It is easy to forget that this is merely an assertion about a 
model, not an assertion about reality [@Bartlett1960].  Only in quantum mechanics do we find 
inherently probabilistic properties: births and deaths of organisms all have
far more mechanistic explanations which we simply summarize in statistical terms:
on average -- that is, summing over a wide range of possible mechanisms we have chosen
not to explicitly model -- an individual death occurs at rate $\lambda$.  
This is in fact an extrinisic factor (variables not modeled explicitly),
which makes birth appear probabilistic; (just as we describe the toss of a coin
as probabilistic when in fact it is a mechanistic outcome of Newton's laws).  
Demographic stochasticity arises only subsequently, when we change scales from
a description at the individual level to one at the population level.

@@

It is possible to simulate such a process exactly using Gillespie's
exact stochastic simulation algorithm [SSA; @Gillespie1977], which
simulates each individual birth or death as a separate event.

We can illustrate both the Gillespie algorithm and the van Kampen
system size expansion using @Levins1969 classic patch model,

@@

where $n$ individuals compete for a
finite number of suitable habitats $N$.  Individuals die a constant
rate $e$, and produce offspring at a constant rate $c$ who then
have a probability of colonizing an open patch that is simply
proportional to the fraction of available patches, $1 - n/N$.  
We choose this model because it has precisely same functional form
as the ubiquitous logistic growth model, $\frac{\ud n}{\ud t} = rn (1 - \tfrac{n}{K})$
but is explicit about how this net growth rate is divided between
birth and death rate.  As we shall see, this is not merely a notational
convience but leads to real differences in observed stochastic properties
of the model.  Though we focus on this simple model for clarity of 
presentation, both computational and mathematical framework presented here 
can easily be applied to alternative models to reach the same conclusions.

Figure \ref{gillespie}
uses this algorithm to compare simulations of @Levins1969 patch model;
(see Appendix A for example code).  In the Levins'  model,

@@

Stochasticity frequently enters our models in other ways as well.
For instance, most population-level models treat all individuals
within the population, or at least within a particular age or stage
class, as identical.  *Individual heterogeneity* in relevant traits
(growth rates, dispersal ability) can be represented probabilistically
as a distribution rather than a fixed value, making it function as
another form of random variation in a model [@Coulson2001;
@Schreiber2010; @Vindenes2015].  *Spatial heterogeneity* can likewise be reflected
in stochastic parameters of a spatially implicit model, or as
spatially varying environmental stochasticity in a spatially explicit
model.  As with other forms of noise [@Melbourne2008; @Bonsall2004],
interaction between individual or spatial variation and other sources
of noise can drive patterns including the potential for persistence,
coexistence, and dispersal of species [@Schreiber2010; @Hart2016;
@Schreiber2017].  *Observational error* acts
much like another form of stochasticity, arising from sampling
effects or limitations of instruments, etc. Yet unlike the sources
described above, this uncertainty does not impact the actual state
of our system, and thus does not give rise to any additional phenomena
or information about that process, though it can obscure it
[@Knape2011; @deValpine2002].
*Deterministic chaos* [@Desharnais2001] is another mechanism of introducing
significant variation into a simple model. Chaos can a look a lot
like noise: indeed, it can be observationally indistinguishable
from a stochastic process [@Ellner2005; @Werndl2009].  Despite
substantial interest in previous decades and the relative ease of
encountering chaos in simple ecological models (particularly in
differences equations), and convincing reconstructions of chaotic
dynamics in laboratory manipulations [@Costantino1997] the prevalence
of chaotic dynamics as a driving factor in most ecological systems
remains unclear. After all, the great variation of natural populations
is hardly surprising: the undeniable influences of variation created
by all those things left out of our models, all that individual
detail from intrinsic, lower level processes and external environmental
processes that becomes aggregated in our simple models all but begs
for a stochastic description.

@@

The loss of a species is without a doubt one of the most important
possible consequences of noise, and unsurprisingly few topics have
recieved greater attention then the question of the persistence of
a given species [e.g. @Kendall1949;  @Bartlett1955; @Bartlett1960;
@Lande1993; @Saether1998; @Coulson2001; @Schreiber2010] and coexistence
of multiple species [@Caswell1978; @Tuljapurkar1980; @Chesson1981;
@Chesson1982; @Chesson1985; @Chesson1989; @Melbourne2007; @Hart2016; @Schreiber2017]
in stochastic models.  Two important trends in this literature have
been establishing existing results in more general and precise language
and theorems [e.g. see @Schreiber2012 and @Schreiber2017 for an excellent
overview and synthesis of persistence and coexistence respectively],
and continuing to extend such results to more complex contexts 
[@Ovaskainen2010 provides one excellent review], in
particular addressing autocorrelated noise [e.g. @Ripa; @Ripa1999; 
@Petchey, @Schreiber2010; @Lee2017; @Spanio2017] , spatial structure
[e.g. @Durrett1994; @Dieckmann2000; @Kerr2002; @Schreiber2010] and stage structure
[e.g. @Cohen1979; @Tuljapurkar1989; @Caswell2009; @Metcalf2015; @Vindenes2017b], 
among other issues.

Demographic noise in particular has fundamental implications for the long-term
equilibrium: any finite population, however big, is guaranteed to
eventually go extinct under demographic noise [e.g. @Kendall1949; @Lande2003].  
In most models it is possible [@Schreiber2006] to define a notion of quasi-stationary 
distribution concentrated around the deterministic attractor,
and to prove the probability of extinction decreases exponentially with 
system size [@Ovaskainen2010; @Schreiber2017]. We will turn our attention
away from the well-trodden ground of persistence and coexistence to focus
on other phenomena that

here we will focus on a few examples of stochastic phenomena 
that can emerge on shorter time scales and provide some intuition for 
how the interaction between stochastic an nonlinear dynamics can 
create these surprising results.
